\section{Introduction}
Modern communication forms such as social media and microblogs are not only rapidly
advancing our understanding of the world but also improving the methods by which we can 
comprehend, and even forecast, the progression of events.
Tracking population-level activities via `massive passive' data has been shown to 
quite accurately shed light into large-scale societal movements. 

Two years back, in KDD 2014, we described EMBERS~\cite{beatingthenews-kdd}, a deployed anticipatory
intelligence system~\cite{bigdata-andy-doyle-embers-paper} that forecasts significant 
societal events (e.g., civil unrest
events such as protests, strikes, and `occupy' events) using a large set of open source
indicators such as news, blogs, tweets, food prices, currency rates, and other public
data. The EMBERS system has been running continuosuly 24x7 for nearly 4 years at this point
and our goal in this paper is to present the discoveries it has enabled,
correct as well as missed
forecasts, and lessons learnt from participating in a forecasting tournament including
our perspectives on the limits of forecasting and ethical considerations. In
particular, we shed insight into the value proposition to an analyst and how EMBERS forecasts
are communicated to its end-users. 

The development of EMBERS is supported by the Intelligence Advanced Research Projects
Activity (IARPA) Open Source Indicators (OSI) program. 
EMBERS forecasts are scored against the GSR (Gold Standard Report), a monthly catalog of 
events as reported in newspapers of record in the above countries. The GSR is compiled by MITRE corporation
using human analysts.
EMBERS currently focuses on multiple regions of the world but for the purpose of this paper
we focus primarily on Latin America, specifically the countries of
Argentina, Brazil, Chile, Colombia, Ecuador, El Salvador, Mexico, Paraguay, Uruguay, and Venezuela.
Similarly, EMBERS generates forecasts for multiple event classes---infleunza like illnesses~\cite{prithwish-ili},
rare diseases~\cite{sdm-saurav}, elections~\cite{aravindan-wei-besc}, domestic political crises~\cite{gdelt-acm-webscience}, and civil unrest---but in this paper we focus primarily on civil unrest as this was the
most challenging event class with hundreds of events every month across the countries studied here.

Our key contributions can be summarized as follows:
\narenc{Refer to sections in each bullet.}
\begin{enumerate}
\item Unlike retrospective studies of predictability, EMBERS forecasts are communicated in real-time before the
event to MITRE/IARPA and scored independently of the authors. 
We present multiple quantiative indicators of EMBERS performance 
as well as insights into how
we made EMBERS forecasts most valuable to analysts. We report two primary
ways in which analysts utilize EMBERS and the use of {\it automated
narratives} to help make EMBERS forecasts as useful as possible.

\item In an attempt to demystify the state-of-the-art
in forecasting and create an open dialogue in the community, we report both successful
forecasts of EMBERS as well as events missed by EMBERS. The events not forecast by EMBERS lead us to
considerations of both limitations of the underlying technology as well as inherent limits to
forecasting large-scale events.
\item While social media is often touted as key to event forecasting
systems such as EMBERS, we present the results of an ablation study to outline
the performance degradation that ensues if data sources 
such as Twitter and Facebook were to be removed from the forecasting pipeline.
\item We consider the separation of civil unrest events into events
that happen with a degree of regularity versus rare
or significant
events, and evaluate the performance of EMBERS in forecasting such
surprising events.
\item We describe our current best understanding of the limitations to
forecasting civil unrest events using technologies like EMBERS and 
also ethical considerations.
\end{enumerate}

